Deciphering enhancer rules underlying early human neuronal and neural crest development.

### 2025-03-21 manuscript.tex
Part of discussion.

>>>>

We built an enhancer-code atlas of early human neuronal and neural crest development. Providing enhancer-codes of
neuronal progenitor DV- and AP-patterning, early neuronal fate acquisition and neural crest induction, migration and
facial mesenchyme specification.\par

In order to create this atlas, we generated a single-cell multiome (combined scATAC-seq and scRNA-seq of the same cell) dataset of the head of a four weeks post conception human embryo and we showed that the embryonic cell states can effectively be reproduced using human neural tube organoids on both transcriptomic level and on the level of the enhancer-codes. Specifically, we identified a total of fifteen distinct cell states with one-to-one matches between the organoid and embryo. The fact that cell states were defined in an unbiased manner by making use of topic modeling and that each of these states were independently identified in two separate datasets increases the confidence of their biological relevance. Furthermore, this high similarity between the organoid and the embyro derived cell states validates neural tube organoids as a biological model for early human neural tube and neural crest development, biological cell states that are otherwise not easily accessible in human.\par

% enhancers

% human tfs paper: https://www.sciencedirect.com/science/article/pii/S0092867418301065?via%3Dihub

% https://www.cell.com/fulltext/s0092-8674(12)01496-1: also about dimerization and that motif "flanks" are important for
% different TFs
% also: https://www.nature.com/articles/s41588-022-01048-5 about flank

%% Enhancers are clusters of TFBS
%%%% https://www.cell.com/cell/fulltext/S0092-8674(16)31251-X
%%%% https://www.cell.com/molecular-cell/fulltext/S1097-2765(22)01215-1
%%%% https://www.sciencedirect.com/science/article/pii/S0959440X21000993

%% Specificity through dimerization
%%%% https://www.cell.com/cell/fulltext/S0092-8674(23)01438-1
%%%% https://pmc.ncbi.nlm.nih.gov/articles/PMC6961570/

% single TF
% https://genome.cshlp.org/content/26/7/882.short

%% pioneer
% https://www.nature.com/articles/s41588-018-0140-x
% https://elifesciences.org/articles/73358
% https://www.cell.com/cell/fulltext/S0092-8674(15)00304-9
% https://www.biorxiv.org/content/10.1101/2024.05.24.595749v3.abstract
% https://www.nature.com/articles/s41576-023-00648-z
% https://www.nature.com/articles/s41586-020-2195-y

%% Low affinity/suboptimization:
%%%% https://www.nature.com/articles/s41586-023-06922-8
%%%% https://www.science.org/doi/10.1126/science.aac6948
%%%% https://pmc.ncbi.nlm.nih.gov/articles/PMC11160683/
%%%% https://www.sciencedirect.com/science/article/abs/pii/S0070215315001842

%% Repression through chromatin closing
% carmen paper & ibrahim paper

%% what about repression through chromatin opening

%% clusters of low affinity HOX binding sites
% https://www.cell.com/cell/fulltext/S0092-8674(14)01518-9
% https://elifesciences.org/articles/28975

% https://www.pnas.org/doi/epub/10.1073/pnas.2411887121
% https://www.nature.com/articles/s41594-024-01235-4

% another paper of face GWAS: https://www.nature.com/articles/s41588-018-0057-4

% might be interesting
% https://www.nature.com/articles/s41594-024-01449-6
% https://www.nature.com/articles/s41586-018-0549-5

% white box models
% DeepNeuralTube
% AP patterning (TP53)
% interpretation of human variation
